---
title: "Project Milestone 4"
authors: "Kosloski & Tonsing"
format: html
editor: visual
---

## Joining data sets

```{r import and clean, include=FALSE}
# Import libraries

library(readr)
library(dplyr)
library(janitor)
library(scales)
library(tidyverse)
library(knitr)
library(kableExtra)
library(tibble)
library(forcats)
library(formattable)
library(plotly)

# Import and clean column names (this was cut and pasted from Milestone2 qmd)

sim_ca <- read_csv("data/sim_novelid_CA.csv", show_col_types = FALSE) |> clean_names()
sim_la <- read_csv("data/sim_novelid_LACounty.csv", show_col_types = FALSE) |> clean_names()
ca_pop <- read_csv("data/ca_pop_2023.csv", show_col_types = FALSE) |> clean_names()

sim_ca <- sim_ca %>% clean_names()
sim_la <- sim_la %>% clean_names()
ca_pop <- ca_pop %>% clean_names()

#clean data (cut and pasted from milestone3 qmd)
sim_la_clean1 <- sim_la %>%
  mutate(county = "Los Angeles") %>%
  rename(
    dt_diagnosis = dt_dx, 
    age_cat = age_category,
    race_ethnicity = race_eth,
    new_infections = dx_new,
    cumulative_infected = infected_cumulative,
    new_unrecovered = unrecovered_new,
    cumulative_unrecovered = unrecovered_cumulative,
    new_severe = severe_new,
    cumulative_severe = severe_cumulative
  ) %>%
  select(
    county, age_cat, sex, race_ethnicity, dt_diagnosis,
    new_infections, cumulative_infected,
    new_unrecovered, cumulative_unrecovered,
    new_severe, cumulative_severe
  )

sim_ca_clean1 <- sim_ca %>%  
  clean_names() %>%
  mutate(
    race_ethnicity = recode(race_ethnicity,
      "1" = "White, Non-Hispanic",
      "2" = "Black, Non-Hispanic",
      "3" = "American Indian or Alaska Native, Non-Hispanic",
      "4" = "Asian, Non-Hispanic",
      "5" = "Native Hawaiian or Pacific Islander, Non-Hispanic",
      "6" = "Multiracial (two or more of above races), Non-Hispanic",
      "7" = "Hispanic (any race)",
      "9" = "Unknown"),
    sex = str_to_title(str_trim(sex)),
    # Remove the word "County" and trim any trailing spaces in county names 
    # to align with population data set
    county = str_remove(county, regex("\\s*County$", ignore_case = TRUE)) %>% str_trim())

sim_ca_clean2 <- sim_ca_clean1 %>%
  select(-time_int)
  
# Clean and standardize LA Morbidity values

sim_la_clean2 <- sim_la_clean1 %>%
  mutate(
    
    # Convert SAS-style date (e.g., 23OCT2023) to YYYY-MM-DD
    dt_diagnosis = as.Date(dt_diagnosis, format = "%d%b%Y"),
    
    race_ethnicity = str_trim(race_ethnicity),
    
    # Ensure consistent capitalization and remove stray spaces in race/ethnicity
    race_ethnicity = str_trim(race_ethnicity),
    race_ethnicity = recode(
      race_ethnicity,
      "Hispanic (Any Race)" = "Hispanic (any race)",
      "American Indian Or Alaska Native, Non-Hispanic" = 
      "American Indian or Alaska Native, Non-Hispanic", 
      "Native Hawaiian Or Pacific Islander, Non-Hispanic" = 
      "Native Hawaiian or Pacific Islander, Non-Hispanic", 
      "Multiracial (Two Or More Of Above Races), Non-Hispanic" =
      "Multiracial (two or more of above races), Non-Hispanic"),
    
    # Clean capitalization for sex variable
    sex = str_to_title(sex),
    
    # Remove the word "County" and trim any trailing spaces in county names 
    # to align with population data set
    county = str_remove(county, regex("\\s*County$", ignore_case = TRUE)) %>% str_trim()
  )

#cleaning of population data set (copied from milestone 3)
ca_pop_clean <- ca_pop %>%
  clean_names() %>%   
  rename(population = pop) %>%
  mutate(
    race_ethnicity = recode(
      race7,
      "WhiteTE NH" = "White, Non-Hispanic",   
      "Black NH"   = "Black, Non-Hispanic",
      "AIAN NH"    = "American Indian or Alaska Native, Non-Hispanic",
      "Asian NH"   = "Asian, Non-Hispanic",
      "NHPI NH"    = "Native Hawaiian or Pacific Islander, Non-Hispanic",
      "MR NH"      = "Multiracial (two or more of above races), Non-Hispanic",
      "Hispanic"   = "Hispanic (any race)"),
    sex    = str_to_title(sex),      # FEMALE → Female
    county = str_to_title(county)    # ALAMEDA COUNTY → Alameda County
  ) %>%
  
  select(county, age_cat, sex, race_ethnicity, population)

#combine 0-17 age froup to match county level data set
ca_pop_collapsed <- ca_pop_clean %>%
  mutate(
    age_cat = case_when(
      age_cat %in% c("0-4", "5-11", "12-17") ~ "0-17",
      TRUE ~ age_cat  # keep other categories unchanged
    )
  ) %>%
  group_by(county, sex, race_ethnicity, age_cat) %>%
  summarise(
    population = sum(population, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r}
# joining morbidity data sets (copied from milestone 3)
morbidity_all <- bind_rows(sim_ca_clean2, sim_la_clean2)

```

```{r aditional cleaning, include=FALSE}
#additional cleaning of combine mortality data sets
morbidity_collapsed<-morbidity_all%>%
  select(-dt_diagnosis)%>%
  group_by(county,sex,race_ethnicity,age_cat)%>%
  summarise(
    new_infections = sum(new_infections, na.rm = TRUE),
    cumulative_infected = sum(cumulative_infected, na.rm = TRUE),
    new_unrecovered = sum(new_unrecovered, na.rm = TRUE),
    cumulative_unrecovered = sum(cumulative_unrecovered, na.rm = TRUE),
    new_severe = sum(new_severe, na.rm = TRUE),
    cumulative_severe = sum(cumulative_severe, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r}
# Join morbidity and population datasets into a new object
ca_id_rate <- full_join(
  morbidity_all,
  ca_pop_collapsed,
  by = c("county", "sex", "race_ethnicity", "age_cat")
)
```

## Visualizations

### Graph

```{r, warning=FALSE, message=FALSE}
ca_id_rate %>%
  group_by(race_ethnicity, sex) %>%
  summarise(rate = (sum(new_infections) / sum(population)) * 100000) %>%
  ggplot(aes(
    x = fct_reorder(race_ethnicity, rate, .fun = max, .desc = FALSE),  # highest at top
    y = rate,
    fill = sex)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(
    title = "Novel Respiratory Disease in California: 
    New Infection Rate per 100,000\nby Race/Ethnicity and Sex, 2023",
    x = "Race/Ethnicity",
    y = "Rate per 100,000",
    fill = "Sex",
    caption = "Source: SIM CA and LA Morbidity Files + CA DOF Population Estimates") +
  theme_minimal() +
  theme(
    plot.title.position = "plot",
    plot.title = element_text(hjust = 0.5))
```

Interpretation: Infection rates for the novel respiratory disease in California show clear racial and ethnic disparities. American Indian or Alaska Native and White, Non-Hispanic populations experience the highest infection rates, followed by Black, Non-Hispanic and Hispanic populations. Native Hawaiian or Pacific Islander, Asian, and Multiracial populations have noticeably lower rates. Differences between males and females within each racial/ethnic group are present but relatively small compared with the larger disparities across groups.

### Table

```{r}
age_table2 <- ca_id_rate %>%
  group_by(age_cat, sex) %>%
  summarise(rate = (sum(new_infections) / sum(population)) * 100000,
            .groups = "drop") %>%
  pivot_wider(names_from = sex, values_from = rate) %>%
  mutate(
    Overall = (Female + Male) / 2
  ) %>%
  arrange(age_cat)

# Define shared palette
pal <- colorRampPalette(c("#FFEA00", "#FFC300","#FFA200","#FF7B00"))(nrow(age_table2))

age_table_color <- age_table2 %>%
  mutate(
    Female = cell_spec(round(Female, 1), "html",
                       background = pal[rank(Female, ties.method = "first")]),
    Male = cell_spec(round(Male, 1), "html",
                     background = pal[rank(Male, ties.method = "first")]),
    Overall = cell_spec(round(Overall, 1), "html",
                        background = pal[rank(Overall, ties.method = "first")])
  )

caption_html <- "<span style='color:black; font-weight:bold; font-size:20px;'>
Novel Respiratory Disease in California: New Infection Rate per 100,000<br>
by Age Group and Sex
</span>"

kable(age_table_color,
      escape = FALSE,
      format = "html",
      caption = caption_html,
      col.names = c("Age Group", "Female", "Male", "Overall")) %>%
  kable_styling(full_width = FALSE)
```

Interpretation: The age groups with the highest infection rate are those over 65 year of age followed by those 18-49 years. Given that the risk does not strictly increase or decrease proportional with age which would be more likely to suggest a biological basis for difference in infection rate, behavioral differences between the age strata should be examined as part of selecting risk reduction strategies.
