---
title: "Team Project: Scenario 1 - Milestone 2"
author: "Kosloski & Tonsing"
format: html
editor: visual
---

## **Description of datasets:**

[Dataset 1: sim_novelid_CA.csv]{.underline}

-   Description - This file has weekly outbreak data for all California counties except Los Angeles. It tracks the number of cases and case severity by age group, sex, race/ethnicity, and county.

-   Relevance - We'll use this data to see how the outbreak is spreading across the state and whether certain groups are being hit harder.

[Dataset 2: sim_novelid_LACounty.csv]{.underline}

-   Description - This file has the same kind of outbreak data but just for Los Angeles County.

-   Relevance - Since LA is left out of the first dataset, this one makes sure we can still look at LA and also compare it to the rest of California.

[Dataset 3: ca_pop_2023.csv]{.underline}

-   Description - This file has 2023 population estimates for every county in California, broken down by age, sex, and race/ethnicity.

-   Relevance - This file provides the "denominators" we need to calculate things like case rates, so we can compare across groups and counties more fairly.

## **Import statements for all datasets:**

We imported the three datasets directly from the class GitHub repository to ensure consistency across teams. We used the readr package to import because it handles CSVs efficiently and allowed us to specify column types if needed. We used the janitor::clean_names() function to standardize column names into snake_case, which makes them easier to work with in R. The import process was documented in our Quarto file so it can be reproduced by others, and we used arguments like show_col_types = FALSE to simplify the output but still keep the option to control column formats and handle missing values later if needed.

**Import Code:**

```{r}

# Import libraries
library(readr)
library(dplyr)
library(janitor)

# Import and clean column names

sim_ca <- read_csv("data/sim_novelid_CA.csv", show_col_types = FALSE) |> clean_names()
sim_la <- read_csv("data/sim_novelid_LACounty.csv", show_col_types = FALSE) |> clean_names()
ca_pop <- read_csv("data/ca_pop_2023.csv", show_col_types = FALSE) |> clean_names()


```

## **Identify data types for 5+ elements/columns/variables:**

We focused on five key variables needed for our analysis: county, age category, sex, race/ethnicity, and new infections. Using glimpse() and summary(), we confirmed that the demographic variables were read in as character strings and will be treated as factors for analysis purposes. The case counts (new_infections) are numeric, which is appropriate for calculating totals and rates. We may also use population from the third dataset as a numeric variable for rate calculations. Going forward, we will likely convert demographic variables to factors to simplify groupings and summaries, and to ensure that dates and counts are consistently numeric/date types across datasets.

```{r}

# Look at structure

glimpse(sim_ca) 
glimpse(sim_la) 
glimpse(ca_pop)
```

## **Provide a basic description of the 5+ data elements:**

```{r by county}
# Character/factor variables: unique categories
unique(sim_ca$sex)
unique(sim_ca$age_cat)
unique(sim_ca$race_ethnicity)
```

```{r LA county}
# Character/factor variables: unique categories
unique(sim_la$sex)
unique(sim_la$age_category)
unique(sim_la$race_eth)
```

```{r whole state}
# Character/factor variables: unique categories
unique(ca_pop$sex)
unique(ca_pop$age_cat)
unique(ca_pop$race7)
```

```{r all except LA}
# Numeric variables: summaries
summary(sim_ca$new_infections)
range(sim_ca$new_infections, na.rm = TRUE)
```

```{r LA county only}
# Numeric variables: summaries
summary(sim_la$dx_new)
range(sim_la$dx_new, na.rm = TRUE)
```

## **Describe cleaning that each data source may need:**

-   Columns in the sim_ca and sim_la data sets are in a different order and will need to be standardized before joining.

-   sim_ca uses the column name dt_diagnosis and sim_la uses dt_dx and the dates are in different formats - This will need to be standardized before joining the data sets.

-   sim_ca uses the column name race_ethnicity and encodes this information as numeric (dbl) values. sim_la uses the column name race_eth and encodes the data as character. ca_pop uses the column name race7 and also encodes the data as character data, but is using different designators for the races. We will need some kind of a "key" for each of the data sets before we can standardize if this is a variable we end up needing.

-   sim_ca uses the column name new_infections and sim_la uses the column name dx_new to encode the daily incidence. These column names will need to be standardized before joining data sets.

-   sim_ca uses the column name cumulative_infected and sim_la uses the column name infected_cumulative to encode the same information. These column names will need to be standardized before joining data sets.

-   sim_ca uses the column name new_unrecovered and sim_la uses the column name unrecovered_new to encode the same information. These column names will need to be standardized before joining data sets.

-   sim_ca uses the column name cumulative_unrecovered and sim_la uses the column name unrecovered_cumulative to encode the same information. These column names will need to be standardized before joining data sets.

-   sim_ca uses the column name new_severe and sim_la uses the column name severe_new to encode the same information. These column names will need to be standardized before joining data sets.

-   sim_ca uses the column name cumulative_severe and sim_la uses the column name severe_cumulative to encode the same information. These column names will need to be standardized before joining data sets.

-   ca_pop has different age categories than both sim_ca and sim_la. The categories will need to be standardized if we chose to use this variable.

-   sim_ca and ca_pop both call the column age_cat, but sim_la calls it age_category. These column names will need to be standardized before joining data sets.
