---
title: "Milestone 6"
author: "Kosloski & Tonsing"
format: html
editor: visual
---

\*\*Everying in italics is cut and paste from the Milestone 6 rubric. My draft responses are non-italicized

***Milestone 6 Rubric (note - Max 1,000 words):***

***Include 2+ visualizations (based on group size): 20pts***

*- 1 (one) print quality table as requested in scenario\
- 1 (one) print quality plot or chart as requested in scenario,\
- 1 (one) additional print quality table or plot of your choice (can support the requested data in the scenario, or answer a different question using the same data sources).\
\
Visualizations should have clear titles, column/axis labels, legends (if applicable), footnotes (if applicable).\
\
Visualizations should also be relevant to the scenario and should support any conclusions drawn in the report.\
\
18 pts for visualizations\
2 pts for visualizations being cohesive (same text formatting, coloring, etc).*

***Problem statement 5pts***

*Include 1-2 paragraphs describing the scenario that your team is aiming to address. This can be similar to the scenario text, but please phrase in your own words.*

For this project, we worked within the scenario of a statewide public health response to a simulated outbreak of a novel respiratory pathogen in California in 2023. The available data included weekly morbidity counts for all counties (with Los Angeles County provided separately) and 2023 population estimates for the same demographic strata. These datasets allowed us to examine disease occurrence by age group, sex, race/ethnicity, and county.

The primary objective of our analysis was to describe how infections were distributed across demographic and geographic groups. By calculating case counts and incidence rates for key populations, we aimed to identify disparities in disease burden and provide evidence to guide resource allocation during the outbreak. This descriptive epidemiologic assessment supports understanding which communities may be most affected and where public health interventions may need to be concentrated.

***Methods 5pts***

*Describe data sources, methods used, strategies for data cleaning, creating new variables, decisions team made about the data, and additional wrangling done for visualizations.*

We used three datasets provided for Scenario 1: weekly morbidity data for all California counties except Los Angeles, a parallel morbidity file for Los Angeles County, and 2023 population estimates for all demographic strata. Together, these sources enabled statewide analysis of new infections and incidence rates by age, sex, race/ethnicity, and county.

All analyses were conducted in R using a workflow centered around the tidyverse. We used readr for data import and dplyr, tidyr, stringr, and forcats for cleaning, recoding, grouping, and reshaping the data. The janitor package standardized column names, and scales and formattable supported numeric formatting. Final tables were produced using knitr and kableExtra, and all figures were created with ggplot2; plotly was used only during development to inspect patterns interactively. Tibbles were used throughout for consistent data handling.

Data cleaning focused on harmonizing demographic variables across files so the datasets could be merged reliably. We aligned variable names, recoded race/ethnicity and age categories to resolve inconsistencies, and ensured county names matched across morbidity and population sources. Records with missing or unusable demographic information were removed. The California and Los Angeles morbidity files were combined into a single statewide dataset and merged with population estimates using shared demographic keys.

To support analysis, we created new variables including total case counts per demographic stratum and incidence rates per 100,000 population. We reordered factor levels and refined category labels to ensure that visualizations were interpretable and aligned with assignment requirements. As a team, we chose to focus on one primary visualization (race/ethnicity by sex) and one primary table (age by sex) to meet the scenario while maintaining clarity. Additional wrangling, such as numeric formatting, ordering categories by rate, and applying consistent labels, was performed to produce publication-ready outputs.

***Results 5pts***

*Describe the results of your analysis, including what is displayed in the visualizations. (Note: no statistical testing is needed)*

Our analysis revealed clear patterns in the distribution of new infections across demographic groups in California. After merging morbidity and population data, we calculated incidence rates per 100,000 population, which were summarized by race/ethnicity and sex in the primary visualization. This horizontal bar chart showed notable disparities: American Indian or Alaska Native and White, Non-Hispanic populations had the highest infection rates, followed by Black and Hispanic populations. Asian, Native Hawaiian or Pacific Islander, and Multiracial groups showed substantially lower rates. Differences between males and females within each racial/ethnic group were relatively small compared with the wider disparities across groups.

We also created a print-quality table summarizing infection rates by age group and sex. Adults aged 65 and older experienced the highest infection rates overall, followed by adults aged 18–49. Rates were lower among those aged 50–64 and lowest among children ages 0–17. The table used color shading to highlight relative magnitudes and clearly demonstrated the gradient of infection burden across age groups. Together, the table and visualization provide a cohesive summary of how both racial/ethnic identity and age shape the distribution of infections in this simulated outbreak.

```{r import and clean, include=FALSE}
# Import libraries

library(readr)
library(dplyr)
library(janitor)
library(scales)
library(tidyverse)
library(knitr)
library(kableExtra)
library(tibble)
library(forcats)
library(formattable)
library(plotly)

# Import and clean column names (this was cut and pasted from Milestone2 qmd)

sim_ca <- read_csv("data/sim_novelid_CA.csv", show_col_types = FALSE) |> clean_names()
sim_la <- read_csv("data/sim_novelid_LACounty.csv", show_col_types = FALSE) |> clean_names()
ca_pop <- read_csv("data/ca_pop_2023.csv", show_col_types = FALSE) |> clean_names()

sim_ca <- sim_ca %>% clean_names()
sim_la <- sim_la %>% clean_names()
ca_pop <- ca_pop %>% clean_names()

#clean data (cut and pasted from milestone3 qmd)
sim_la_clean1 <- sim_la %>%
  mutate(county = "Los Angeles") %>%
  rename(
    dt_diagnosis = dt_dx, 
    age_cat = age_category,
    race_ethnicity = race_eth,
    new_infections = dx_new,
    cumulative_infected = infected_cumulative,
    new_unrecovered = unrecovered_new,
    cumulative_unrecovered = unrecovered_cumulative,
    new_severe = severe_new,
    cumulative_severe = severe_cumulative
  ) %>%
  select(
    county, age_cat, sex, race_ethnicity, dt_diagnosis,
    new_infections, cumulative_infected,
    new_unrecovered, cumulative_unrecovered,
    new_severe, cumulative_severe
  )

sim_ca_clean1 <- sim_ca %>%  
  clean_names() %>%
  mutate(
    race_ethnicity = recode(race_ethnicity,
      "1" = "White, Non-Hispanic",
      "2" = "Black, Non-Hispanic",
      "3" = "American Indian or Alaska Native, Non-Hispanic",
      "4" = "Asian, Non-Hispanic",
      "5" = "Native Hawaiian or Pacific Islander, Non-Hispanic",
      "6" = "Multiracial (two or more of above races), Non-Hispanic",
      "7" = "Hispanic (any race)",
      "9" = "Unknown"),
    sex = str_to_title(str_trim(sex)),
    # Remove the word "County" and trim any trailing spaces in county names 
    # to align with population data set
    county = str_remove(county, regex("\\s*County$", ignore_case = TRUE)) %>% str_trim())

sim_ca_clean2 <- sim_ca_clean1 %>%
  select(-time_int)
  
# Clean and standardize LA Morbidity values

sim_la_clean2 <- sim_la_clean1 %>%
  mutate(
    
    # Convert SAS-style date (e.g., 23OCT2023) to YYYY-MM-DD
    dt_diagnosis = as.Date(dt_diagnosis, format = "%d%b%Y"),
    
    race_ethnicity = str_trim(race_ethnicity),
    
    # Ensure consistent capitalization and remove stray spaces in race/ethnicity
    race_ethnicity = str_trim(race_ethnicity),
    race_ethnicity = recode(
      race_ethnicity,
      "Hispanic (Any Race)" = "Hispanic (any race)",
      "American Indian Or Alaska Native, Non-Hispanic" = 
      "American Indian or Alaska Native, Non-Hispanic", 
      "Native Hawaiian Or Pacific Islander, Non-Hispanic" = 
      "Native Hawaiian or Pacific Islander, Non-Hispanic", 
      "Multiracial (Two Or More Of Above Races), Non-Hispanic" =
      "Multiracial (two or more of above races), Non-Hispanic"),
    
    # Clean capitalization for sex variable
    sex = str_to_title(sex),
    
    # Remove the word "County" and trim any trailing spaces in county names 
    # to align with population data set
    county = str_remove(county, regex("\\s*County$", ignore_case = TRUE)) %>% str_trim()
  )

#cleaning of population data set (copied from milestone 3)
ca_pop_clean <- ca_pop %>%
  clean_names() %>%   
  rename(population = pop) %>%
  mutate(
    race_ethnicity = recode(
      race7,
      "WhiteTE NH" = "White, Non-Hispanic",   
      "Black NH"   = "Black, Non-Hispanic",
      "AIAN NH"    = "American Indian or Alaska Native, Non-Hispanic",
      "Asian NH"   = "Asian, Non-Hispanic",
      "NHPI NH"    = "Native Hawaiian or Pacific Islander, Non-Hispanic",
      "MR NH"      = "Multiracial (two or more of above races), Non-Hispanic",
      "Hispanic"   = "Hispanic (any race)"),
    sex    = str_to_title(sex),      # FEMALE → Female
    county = str_to_title(county)    # ALAMEDA COUNTY → Alameda County
  ) %>%
  
  select(county, age_cat, sex, race_ethnicity, population)

#combine 0-17 age froup to match county level data set
ca_pop_collapsed <- ca_pop_clean %>%
  mutate(
    age_cat = case_when(
      age_cat %in% c("0-4", "5-11", "12-17") ~ "0-17",
      TRUE ~ age_cat  # keep other categories unchanged
    )
  ) %>%
  group_by(county, sex, race_ethnicity, age_cat) %>%
  summarise(
    population = sum(population, na.rm = TRUE),
    .groups = "drop"
  )
```
```{r, include=FALSE}

# joining morbidity data sets (copied from milestone 3)
morbidity_all <- bind_rows(sim_ca_clean2, sim_la_clean2)
```
```{r aditional cleaning, include=FALSE}
#additional cleaning of combine mortality data sets
morbidity_collapsed<-morbidity_all%>%
  select(-dt_diagnosis)%>%
  group_by(county,sex,race_ethnicity,age_cat)%>%
  summarise(
    new_infections = sum(new_infections, na.rm = TRUE),
    cumulative_infected = sum(cumulative_infected, na.rm = TRUE),
    new_unrecovered = sum(new_unrecovered, na.rm = TRUE),
    cumulative_unrecovered = sum(cumulative_unrecovered, na.rm = TRUE),
    new_severe = sum(new_severe, na.rm = TRUE),
    cumulative_severe = sum(cumulative_severe, na.rm = TRUE),
    .groups = "drop"
  )
```
```{r, include=FALSE}
# Join morbidity and population datasets into a new object
ca_id_rate <- full_join(
  morbidity_all,
  ca_pop_collapsed,
  by = c("county", "sex", "race_ethnicity", "age_cat")
)
```
```{r, echo=FALSE, message=FALSE, warning=FALSE}
ca_id_rate %>%
  group_by(race_ethnicity, sex) %>%
  summarise(rate = (sum(new_infections) / sum(population)) * 100000) %>%
  ggplot(aes(
    x = fct_reorder(race_ethnicity, rate, .fun = max, .desc = FALSE),  # highest at top
    y = rate,
    fill = sex)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(
    title = "Novel Respiratory Disease in California: 
    New Infection Rate per 100,000\nby Race/Ethnicity and Sex, 2023",
    x = "Race/Ethnicity",
    y = "Rate per 100,000",
    fill = "Sex",
    caption = "Source: SIM CA and LA Morbidity Files + CA DOF Population Estimates") +
  theme_minimal() +
  theme(
    plot.title.position = "plot",
    plot.title = element_text(hjust = 0.5))
```
Interpretation: Infection rates for the novel respiratory disease in California show clear racial and ethnic disparities. American Indian or Alaska Native and White, Non-Hispanic populations experience the highest infection rates, followed by Black, Non-Hispanic and Hispanic populations. Native Hawaiian or Pacific Islander, Asian, and Multiracial populations have noticeably lower rates. Differences between males and females within each racial/ethnic group are present but relatively small compared with the larger disparities across groups.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
age_table2 <- ca_id_rate %>%
  group_by(age_cat, sex) %>%
  summarise(rate = (sum(new_infections) / sum(population)) * 100000,
            .groups = "drop") %>%
  pivot_wider(names_from = sex, values_from = rate) %>%
  mutate(
    Overall = (Female + Male) / 2
  ) %>%
  arrange(age_cat)

# Define shared palette
pal <- colorRampPalette(c("#FFEA00", "#FFC300","#FFA200","#FF7B00"))(nrow(age_table2))

age_table_color <- age_table2 %>%
  mutate(
    Female = cell_spec(round(Female, 1), "html",
                       background = pal[rank(Female, ties.method = "first")]),
    Male = cell_spec(round(Male, 1), "html",
                     background = pal[rank(Male, ties.method = "first")]),
    Overall = cell_spec(round(Overall, 1), "html",
                        background = pal[rank(Overall, ties.method = "first")])
  )

caption_html <- "<span style='color:black; font-weight:bold; font-size:20px;'>
Novel Respiratory Disease in California: New Infection Rate per 100,000<br>
by Age Group and Sex
</span>"

kable(age_table_color,
      escape = FALSE,
      format = "html",
      caption = caption_html,
      col.names = c("Age Group", "Female", "Male", "Overall")) %>%
  kable_styling(full_width = FALSE)
```
Interpretation: The age groups with the highest infection rate are those over 65 year of age followed by those 18-49 years. Given that the risk does not strictly increase or decrease proportional with age which would be more likely to suggest a biological basis for difference in infection rate, behavioral differences between the age strata should be examined as part of selecting risk reduction strategies.

***Discussion 5pts***

*Provide interpretation of results and what they tell your team about the given scenario.*

The findings from this project indicate that the burden of the simulated respiratory outbreak was not evenly distributed across California’s population. The race/ethnicity–sex visualization highlights pronounced disparities, with certain racial and ethnic groups experiencing significantly higher infection rates than others. These patterns likely reflect a combination of structural, occupational, and environmental factors rather than biological differences alone. The minimal differences by sex support the interpretation that social and structural factors, access to preventative resources, crowded living or working conditions, or county-level transmission dynamics, play a larger role than sex in shaping infection risk.

The age-specific results further illuminate these differences. Older adults had markedly higher infection rates, consistent with both greater biological susceptibility and potential differences in exposure or health-seeking behavior. The substantial burden among adults aged 18–49 suggests that working-age individuals may represent an important driver of community transmission, potentially due to workplace exposures or caregiving responsibilities.

Taken together, these results indicate that an effective outbreak response would require targeted outreach to the most affected racial/ethnic groups and increased support for older adults. Enhancing access to testing, treatment, and prevention resources, along with tailored communication strategies, would likely help reduce both overall transmission and the disparities observed in this simulated scenario. These findings illustrate the importance of demographic-specific analyses in guiding equitable public health action.

***Uniform theme throughout document 10pts***

*Final html output should all be uniform in format, coloring, text style (it should not look like multiple people have contributed to it)*

***Html output published on RPubs 10pts***

*Final outputs should include all necessary code to generate html output. No excess codes or unnecessary comments should be included. For html, make sure no code is printed so that it looks like a polished report.*
